---
title: "Aqua sync isd update"
author: "Justin Pomeranz"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# `isdbayes` estimates  

## Setup
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(tidybayes)
library(isdbayes)

dat <- readRDS("derived_data/filtered_size_2024-02-07.RDS")
dat <- dat %>%
  mutate(group_id = as.character(group_id))

posts <- readRDS("results/brm_posts_2024-03-05.rds")
sample_gm <- readRDS("results/sample_gm.rds")

```


## Load model fits

## Posterior draws  


```{r}
# CI for posterior draws ####
head(posts)
```

## Credible Intervals of Posterior draws  


```{r, warning=FALSE, message=FALSE}
ci_posts <- posts %>%
  group_by(group_id,
           dat_id,
           geographical_latitude,
           organism_groups,
           n_obs_dat,
           xmin, 
           xmax) %>%
  summarize(
    median = median(.epred),
    l95 = quantile(.epred, probs = 0.025),
    u95 = quantile(.epred,  probs = 0.975)) %>%
  mutate(width = abs(l95 - u95),
         size_range = log10(xmax) - log10(xmin))

ci_posts
```


### How many "sites" with arbitrary CrI widths?  
```{r}
# ci_posts widths ####
ci_posts %>%
  ungroup() %>%
  select(group_id, dat_id, width) %>%
  filter(width <=2) %>%
  nrow()
# 774 groups

ci_posts %>%
  ungroup() %>%
  select(group_id, dat_id, width) %>%
  filter(width <=1) %>%
  nrow()

# 620

ci_posts %>%
  ungroup() %>%
  select(group_id, dat_id, width) %>%
  filter(width <=0.5) %>%
  nrow()

# 558

ci_posts %>%
  ungroup() %>%
  select(group_id, dat_id, width) %>%
  filter(width <0.96) %>%
  nrow()

# 615 --> matches MLE count with vec diff 0.5
```

## PLots of CrI across latitudes  

```{r, warning=FALSE, message=FALSE}
ci_posts %>%
  ggplot(aes(x = abs(geographical_latitude),
             y = median,
             ymin = l95,
             ymax = u95,
             group = group_id,
             color = organism_groups)) +
  geom_pointrange() +
  labs(title = "All estimates and 95% CrI") +
  #scale_x_log10() +
  #facet_wrap(~organism_groups, scales = "free") +
  NULL

ci_posts %>%
  ggplot(aes(x = abs(geographical_latitude),
             y = median,
             ymin = l95,
             ymax = u95,
             group = group_id,
             color = organism_groups)) +
  geom_pointrange() +
  labs(title = "All estimates and 95% CrI; facetted by organism groups") +
  #scale_x_log10() +
  facet_wrap(~organism_groups, scales = "free") +
  theme(legend.position = "none") +
  NULL

ci_posts %>%
  filter(width <1) %>%
  ggplot(aes(x = abs(geographical_latitude),
             y = median,
             ymin = l95,
             ymax = u95,
             group = group_id,
             color = organism_groups)) +
  geom_pointrange() +
  labs(title = "Estimates with 95% CrI < 1") +
  #scale_x_log10() +
  #facet_wrap(~organism_groups, scales = "free") +
  NULL

```


# Bayesian Summary Statistics  

```{r warning=FALSE, message=FALSE}
# sample yrep from posterior
n_pred = 500
post_gm <- posts %>%
  filter(.draw <= 100) %>%
  select(xmin, xmax, group_id, .epred, .draw) %>%
  expand_grid(individual = 1:n_pred) %>% 
  ungroup()  %>% 
  mutate(
    yrep = rparetocounts(nrow(.),
                         lambda = .epred,
                         xmin = xmin,
                         xmax = xmax))

reorder_ids = sample_gm %>% 
  ungroup() %>% 
  #group_by(group_id) %>% 
  #reframe(rank_gm = exp(mean(log(yrep)))) %>% 
  arrange(gm) %>% 
  mutate(sample_id_ordered = 1:nrow(.)) %>%
  select(group_id, sample_id_ordered)

post_gm_summary <- post_gm %>%
  group_by(group_id, xmin, xmax, .draw) %>%
  reframe(gm = exp(mean(log(yrep))),
          median_dw = median(yrep)) %>% 
  left_join(reorder_ids)

```

## Posterior Prediction of Geometric Mean body size  

```{r warning=FALSE, message=FALSE}
post_gm_summary %>% 
  ggplot(aes(x = sample_id_ordered, y = gm)) +
  stat_pointinterval(alpha = 0.5,
                     size = 0.2,
                     aes(color = "Posterior Predictive",
                         shape = "Posterior Predictive")) +
  scale_y_log10() +
  geom_point(data = sample_gm %>% 
               left_join(reorder_ids), 
             aes(y = gm, color = "Raw Data", shape = "Raw Data"),
             size = 1) +
  guides(fill = "none") +
  guides(shape = "none") +
  scale_color_manual(values = c("#56b4e9", "black")) +
  labs(x = "Rank GM body size",
       y = "Geometric mean body size (mgDM)") +
  guides(color = guide_legend(override.aes = list(shape = c(16, 17)),
                              title = "")) +
  #theme(legend.position = c(0.2, 0.95)) +
  NULL

post_gm_summary %>% 
  left_join(dat %>%
              distinct(group_id,
                       dat_id,
                       geographical_latitude,
                       organism_groups)) %>%
  ggplot(aes(x = sample_id_ordered, y = gm)) +
  stat_pointinterval(alpha = 0.5,
                     size = 0.2,
                     aes(color = "Posterior Predictive",
                         shape = "Posterior Predictive")) +
  scale_y_log10() +
  geom_point(data = sample_gm %>% 
               left_join(reorder_ids) %>%
               left_join(dat %>%
                           distinct(group_id,
                                    dat_id,
                                    geographical_latitude,
                                    organism_groups)), 
             aes(y = gm,
                 color = organism_groups,
                 shape = "Raw Data"),
             size = 1) +
  guides(fill = "none") +
  guides(shape = "none") +
  #scale_color_manual(values = c("#56b4e9", "black")) +
  labs(x = "Rank GM body size",
       y = "Geometric mean body size (mgDM)") +
  # guides(color = guide_legend(override.aes = list(shape = c(16, 17)),
  #                            title = "")) +
  facet_wrap(~organism_groups,
             scales = "free") +
  #theme(legend.position = c(0.2, 0.95)) +
  NULL
```


## Explore random subset of "wide" fish fits  

```{r warning=FALSE, message=FALSE}
wide_fish_groups <- post_gm_summary %>% 
  left_join(dat %>%
              distinct(group_id,
                       dat_id,
                       organism_groups)) %>%
  filter(organism_groups == "Fish",
         sample_id_ordered > 1500) %>%
  sample_n(100) %>% 
  pull(group_id)

post_gm_summary %>% 
  filter(group_id %in% wide_fish_groups) %>%
  ggplot(aes(x = sample_id_ordered, y = gm)) +
  stat_pointinterval(alpha = 0.5,
                     size = 0.2,
                     aes(color = "Posterior Predictive",
                         shape = "Posterior Predictive")) +
  scale_y_log10() +
  geom_point(data = sample_gm %>% 
               filter(group_id %in% wide_fish_groups) %>%
               left_join(reorder_ids), 
             aes(y = gm, color = "Raw Data", shape = "Raw Data"),
             size = 1) +
  guides(fill = "none") +
  guides(shape = "none") +
  scale_color_manual(values = c("#56b4e9", "black"))

posts %>% 
  filter(group_id %in% wide_fish_groups) %>%
  ggplot(aes(x = abs(geographical_latitude),
             group = group_id,
             y = .epred)) +
  stat_pointinterval(alpha = 0.5,
                     size = 0.2) +
  labs(y = "lambda; 95% CrI")
```

# MLE Count  

```{r}
mle_count <- readRDS("results/mle_count_vecDiff-0.5_2024-03-01.rds")

mle_count %>%
  filter(!is.na(mle_estimate)) %>%
  nrow()

mle_count %>%
  filter(!is.na(mle_estimate)) %>%
  nrow() / nrow(mle_count)
```

Only 615 ~ 24% of data has estimates
When I change the `bvec = ` argument, we get more, but increase in CI widths. 

```{r warning=FALSE, message=FALSE}

mle_count$group_id <- as.character(mle_count$group_id)
mle_lat <- left_join(
  mle_count,
  dat %>%
    distinct(group_id,
             dat_id,
             geographical_latitude,
             organism_groups))

ggplot(mle_lat,
       aes(
         x = abs(geographical_latitude),
         group = group_id,
         color = organism_groups,
         y = mle_estimate,
         ymin = conf_lo,
         ymax = conf_hi,
         )) +
  geom_pointrange() +
  labs(y = "lambda",
       x = "Absolute latitude",
       title = "MLECount estimate")
```

